---
title: "Zero inflated Poisson distribution"
author: "For Master 2 Math SV"
date: " January2024"
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1.  The data

We study  the abundance of fish species at $n = 89$ sites in the Barents Sea (@Fossheim). The data
are available in the file BarentsFish.csv where the first 4 columns correspond to four environmental covariates
covariates (latitude, longitude, depth, temperature) and the next 30 columns are the abundances of 30
species.

```{r data}
abundance <- read.csv("BarentsFish.csv", sep=";")
View(abundance)
```

In the following, we will consider only one fish species, for example the 20th ('Se_ma = Sebastes marinus =
Golden redfish) and we will note $1 \leq  i \leq n$.
$$Y_i = \mbox{ abundance of golden redfish in station i}.$$

   **1.** Explore the data with standard tools (means, histograms...)

```{r data hist}
library(ggplot2)
ggplot(abundance,aes(Se_ma))+geom_histogram()
```


Observe an over-representation of null values. We propose to modelize this over inflation of $0$. 


### 2.   Zero-inflated Poisson model

We propose to consider the following Zero Inflation Poisson distribution (ZIP) Let $Z_i$ be a latent variable such that 
$$Z_i \sim_{i.i.d} \mathcal{B}ern(1-\pi)$$
Then

\begin{equation}
Y_i | Z_i \sim (1-Z_i) \delta_{\{0\}} + Z_i \mathcal{P}(\mu_i)
\end{equation}

where $\mathcal{P}$ is the Poisson distribution. 

 2.   Write the marginal distribution of $Y_i$
 
 3. Derive $\mathbb{E}[Y_i]$ and $P(Y_i=0)$
 
 4.  Write the complete log likelihood $\log p_\theta(\mathbf{Y}, \mathbf{Z})$ of the model where $\theta= (\pi,\mu)$.
 
 We propose to maximize likelihood with respect to the  parameters using the EM algorithm 
 
 5.  Write the  corresponding E-step.
 
 6.  Write the  corresponding M-step.

 7.  Suggest an initial value for the parameter $\theta$.

 8.  Code the EM algorithm.


### 3. ZIP with covariates 
We now consider a model similar to ZIP but taking into account the environmental covariates.
We note $x_i$ the vector comprising these covariates for the site $i$:
$$x_i = [1, \mbox{latitude}_i, \mbox{longitude}_i, \mbox{depth}_i, \mbox{temperature}_i ].$$
We therefore pose : $(Z_i)_{1 ,\leq i \leq n}$ independent,$(Y_i | Z_i)_{1 ,\leq i \leq n}$ independent
and
\begin{equation}
\begin{array}{cclllcl}
Z_i &\sim& \mathcal{B}\mbox{ern}(\pi_i) &   \mbox{with} & \log \left(\frac{\pi_i}{1- \pi_i}\right) &=& x_i^T \alpha\\
Y_i | Z_i &\sim& (1-Z_i) \delta_{\{0\}} + Z_i \mathcal{P}(\mu_i)& \mbox{with} & \log \mu_i &=& x_i^T \beta
\end{array}
\end{equation}

The vectors $\alpha$  and $\beta$ contain the regression coefficients to predict absence and abon-
dance conditional on the presence of the species at each site.
      
   9. Write the full log likelihood $p_{\theta}(\mathbf{Y}, \mathbf{Z})$  of this new model as a function of the parameter $\theta = (\alpha, \beta)$.
      
   10. Write the E-step.
   
   11. Write the M-step.
      
   12. Propose an initial value for the parameter $\theta$.
      
   13. Code the EM algorithm


### References

